---
title: "Calorie Burning Report"
output: pdf_document
---

Our [R Markdown](http://rmarkdown.rstudio.com) Notebook which will form the final report. When you execute code within the notebook, the results appear beneath the code.

You can run code here and insert results directly.

```{r}
options(digits=4)
muscledata = read.table("muscle-incomplete.txt", header=T, na.strings = "NA")
muscleSum = summary(muscledata)
```
Here are some descriptive statistics.
```{r}
plot(muscledata)
```

Let's check correlation between the weight of an individual and the intensity of their workout, as defined by the calories burnt per hour, using the Pearson correlation coefficient:
```{r}
cor.test(muscledata$weight, muscledata$calhour, alternative="two.sided", method="pearson")
```

The P value is way, way more than 0.05. We accept the null hypothesis, there seems to be no significant correlation.

Let's try plotting a simple linear model to this relationship.
```{r}
weightCalhr = lm(weight~calhour, data=muscledata)
summary(weightCalhr)
```

The linear model confirms our suspicion that there is no relationship here.