---
title: "Calorie Burning Report"
output: pdf_document
---

# Methods and procedure

## Data exploration
First let's load the data and look at the summary.
```{r}
requirements = c("nlme", "effects", "pastecs", "lattice", "psych", "ggplot2", "GGally", "mice", "VIM", "aod", "BaM", "lme4")
if (length(setdiff(requirements, rownames(installed.packages()))) > 0) {
  invisible(install.packages(setdiff(requirements, rownames(installed.packages())))) 
}
for (i in seq(1, length(requirements))) {
  invisible(library(requirements[i], character.only=T))
}
options(digits=4)
muscledata = read.table("muscle-incomplete.txt", header=T, na.strings = "NA")
summary(muscledata)
```

Here are some descriptive statistics.
```{r}
plot(muscledata)
```

Let's check correlation between the weight of an individual and the intensity of their workout, as defined by the calories burnt per hour, using the Pearson correlation coefficient:
```{r}
cor.test(muscledata$weight, muscledata$calhour, alternative="two.sided", method="pearson")
```

The P value is way, way more than 0.05. We accept the null hypothesis, there seems to be no significant correlation.

Let's try plotting a simple linear model to this relationship.
```{r}
weightCalhr = lm(weight~calhour, data=muscledata)
summary(weightCalhr)
```
The linear model confirms our suspicion that there is no relationship here.

## Missing data exploration

```{r}
aggr(muscledata, numbers = TRUE, prop = FALSE, ylab = c("Histogram of missing data", "Pattern"))
```


```{r}
aggr(muscledata, combined=TRUE, numbers = TRUE, prop = TRUE, cex.numbers=0.87, varheight = FALSE)
barMiss(muscledata[,c("calhour","weight")])
barMiss(muscledata[,c("calories","weight")])
histMiss(muscledata)
```



## Complete case analysis

## Multiple imputation analysis

```{r}
invisible(capture.output(muscledata.imp <- mice(muscledata, meth = c("", "", "pmm"), m=100)))
invisible(muscledata.fit <- with(data=muscledata.imp, exp=glm(calories~weight+calhour+weight*calhour)))
MI.matrix<-matrix(0,100,4)
for(k in 1:100){ 
  invisible(MI.matrix[k,]<-coefficients(muscledata.fit$analyses[[k]]))
  }
MI.results=data.frame(Intercept=MI.matrix[,1], weight=MI.matrix[,2],calhour=MI.matrix[,3], interaction=MI.matrix[,4])
head(MI.results[1:10,])
muscledata.est <- pool(muscledata.fit)
summary(muscledata.est)
```


## IPW analysis

```{r}
muscledata$r<-as.numeric(!is.na(muscledata$calories))
head(muscledata)
muscledata.ipw.glm<-glm(r ~ weight+calhour+weight*calhour, data=muscledata,family=binomial)
summary(muscledata.ipw.glm)
```

```{r}
muscledata$w<-1/fitted(muscledata.ipw.glm)
muscledata.results.ipw<- glm(calories~weight+calhour+weight*calhour, data=muscledata, weights=muscledata$w)
summary(muscledata.results.ipw)
```



# Discussion

# Conclusion


